Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-25 paper_v1.Rnw
Added open response analysis to study 2.

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller but more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. Identifying learners who were likely to drop out was achieved with machine learning.

\begin{description}
  \item[H4a] Learners who report a `lack of time' can be divided into those with too many higher priority obligations (work, school, family) and those with low volition.
  \item[H4b] Learners who do not report specific obligations (e.g., work, school, family) tend to fall in the low volition than in the high-priority obligations group, and vice versa.
\end{description}

@@

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A very low response rate was expected, given that this subpopulation was defined by low engagement. 756 out of 6,050 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated using chained equations based on responses to all survey questions\footnote{reference to MICE}. All estimates were pooled across ten imputations, with standard errors adjusted for added variance. 85\% of respondents were male, 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.

@@

Two research assistants independently developed a codebook for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Mechanical Turk, where 250 randomly selected responses were each coded by four `classification experts'. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. The updated codebook was then applied by Mechanical Turk `classification experts' to the 198 remaining open responses (four coders per response).

Besides these open responses, learners reported the difficulty of the course material and the extent to which their progress was hindered by a number of obstacles (see Table [XX]). The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and one item on volition (about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

Reasons for learner attrition were measured using open response as well as multiple-choice questions. As we were interested in learners who left the course unsatisfied, we focused on the subpopulation that reported being only somewhat or less satisfied with their progress in the course (73\% of respondents). Almost half (47\%) the respondents found the course somewhat, very, or extremely difficult. When asked about the extent their progress was hindered by certain challenges, 18\% reported problems with computer or Internet access, 8\% reported technical difficulties, 17\% had difficulties with their level of English, and 24\% found the course requirements unclear. While two in three (66\%) respondents were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. volition). The two measures, which only differ in the specified level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.

@@

Hypotheses H1, H2, and H3 about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress. The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction. The results fully support $H1$, and partially support $H2$ and $H3$.

\begin{table*}[ht]
\caption{Descriptive and inferrential statistics for psychological measures by learner success}
\label{tab:psych}
\small
\center
\begin{tabular}{lcccc}
\toprule
 & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
\midrule
\emph{Relative Progress} &  &  &  \\
\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
 &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
 \emph{Satisfaction} &  &  &  \\
\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
\bottomrule
% \multicolumn{4}{l}{Two-sided $t$-test: $^{***} < .001$, $^{**} < .01$, $^{*} < .05$} \\
\end{tabular}
\end{table*}

>>>>

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. We developed a predictive model to identify learners who were likely to drop out and ask them to complete a survey.

\begin{description}
  \item[H4a] Learners who report having `not enough time' can be divided into those with too many higher priority obligations (work, school, family) and those with low volition.
  \item[H4b] Learners who do not report specific obligations (e.g., work, school, family) tend to fall in the low volition than in the high-priority obligations group, and vice versa.
\end{description}

@@

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for a subpopulation that was defined by low engagement. And yet, 756 out of 6,050 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated by predictive mean matching using responses to all survey questions\footnote{The $R$ mice package was used to perform multiple imputation and subsequent analyses.}. All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.

@@

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Mechanical Turk in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts'. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see [URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

Besides these open responses, learners rated their relative progress and satisfaction with their progress in the course, which served as measures of personal success. They also rated the extent to which their progress was hindered by a number of obstacles. The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and one item on volition (about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

The majority of survey respondents (73\%; 550 of 756) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. volition). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.

Reasons for learner attrition were measured using open response as well as multiple-choice questions. Based on learners' coded open responses ($n=399$), 84\% had `not enough time for the course', 18\% indicated that `the course format or teaching style was not a good fit', 10\% expressed that `the level of the course was too advanced', 7\% found that `the course was not interesting or valuable enough', 4\% expressed a need for `additional course materials to learn the topic', 2\% experienced `technical difficulties', and 1\% reported facing a `language barrier'. A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).

To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time, which revealed that about half of them (53\%) did not specify a particular committment. This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volition, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by committments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses 4a and 4b: while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volition and reporting no specific reasons for having not enough time was indicative thereof.

@@

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress. The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction. The results fully support $H1$, and partially support $H2$ and $H3$.

\begin{table*}[ht]
\caption{Descriptive and inferrential statistics for psychological measures by learner success}
\label{tab:psych}
\small
\center
\begin{tabular}{lcccc}
\toprule
 & $N$ & Goal Striving ($H1$) & Social Belonging ($H2$) & Growth Mindset ($H3$) \\
\midrule
\emph{Relative Progress} &  &  &  \\
\quad Successful & 241 & 3.46 $(SD=0.91)$ & 4.70 $(SD=0.74)$ &  4.58 $(SD=0.99)$ \\
\quad Unsuccessful & 515 & 2.96 $(SD=0.93)$ & 4.63 $(SD=0.70)$ & 4.38 $(SD=0.87)$ \\
 &  & $t_{201}=6.3, p<.001, d=0.55$ & $t_{60}=1.0, p>.25, d=.09$ & $t_{152}=2.4, p=.017, d=0.21$ \\
 \emph{Satisfaction} &  &  &  \\
\quad Successful & 325 & 3.43 $(SD=0.83)$ & 4.76 $(SD=0.70)$ & 4.47 $(SD=0.93)$ \\
\quad Unsuccessful & 431 & 2.88 $(SD=0.97)$ & 4.58 $(SD=0.71)$ & 4.42 $(SD=0.91)$ \\
&  & $t_{166}=7.2, p<.001, d=0.59$ & $t_{158}=3.1, p=.002, d=.26$ & $t_{175}=0.6, p>.25, d=.05$ \\
\bottomrule
\end{tabular}
\end{table*}
